---
title: "WHO Moratlity Data"
author: "S Bashir"
date: "9 October 2016"
output: html_document
---

In this document, the steps to *tidy* the [World Health Organisation's (WHO)](www.who.int) mortality data for use in R are documented. The following steps are presented.

* Data source: <http://www.who.int/healthinfo/mortality_data/en/>
  + It is assumed that data has been downloaded
* Reading in the WHO mortality data
* Reading the International Classificaion of Diseases data
* Keep the **neoplasm**, **melanoma**, **lymphoma** and **leukaemia** data for ICD10 part 1
  + Tranform the data into long format for the number of deaths
* Transform the population data in long format
* Merge the long population data to the long neoplasm data
  + Care has to be taken with respect to the format of the age groups. 
  + Combine the youngest age groups into a new 0-4 years. 
  + Consider adding an **all** age group category (**not urgent**)
* Calculate the mortality rate per 100,000 population in a new variable
  + rate = (100000*deaths)/pop
* Write a function that will extract data based on country, cause of death, year and sex
  + If any of these are left blank then all the data is extracted
  + A data frame containing the requested data should be returned.
* Next steps will defined later
  * Sumamry graphs
  * Summary tables
  
**Note** There is no not need to use *R Shiny* for this exercise. Only program in this R markdown document.

## Data source

The data was downloaded from <http://www.who.int/healthinfo/mortality_data/en/>. It is assumed that data has been downloaded and resides in a directory "OrigData"

## Reading in the WHO mortality data

The following code will automatically read in the WHO mortality data.
```{r ReadWHOdata, echo=FALSE}
# Load libraries
library(stringr)
library(readr)
library(tibble)

# Set the directory where you have saved the data. 
# This is not a great solution but it works and is easier for the moment, Later it should change.
setwd("OrigData/")

# Select the data files (exclude .zip, .doc, .xls)
csvWHO <- list.files(path=".", full.names = FALSE) %>%
          str_subset("^((?!zip|doc|xls).)*$")

# Use the lapply to read all the data into one list
dataList <- lapply(csvWHO, read_csv, guess_max = min(300000, Inf))

# Extract the data into data.frames (tbl) but start by giving them the dataest you want.
# I convert them to lower case as they are not consistent in the original.
names(dataList) <- str_to_lower(csvWHO) 
list2env(dataList, envir=.GlobalEnv)

# dataList is not longer needed so we will remove it.
rm(dataList)
# Removing the ICD 7, 8, 9 and 10 part 2 data as we will not use them for the moment.
# Also we do not want to have problems with the memory 
rm(morticd7, morticd8, morticd9, morticd10_part2)

# Some garbage collection to free up the memory (although this is something that is
# automatically by kernel). In future we may not need this step.
gc()
```

## Reading the International Classificaion of Diseases data

**Joana:** Please read in the data form the word document here.

```{r readICDdata, echo=TRUE}
library(xml2)
library(tidyr)
library(dplyr)
library(XML)
library(rvest)

# Getting icd10 codes from the website www.wolfbane.com
# 1st URL: 4-digit codes
# 2nd URL: letter-plus-2or3digit codes
URL <- c('http://www.wolfbane.com/icd/icd10a1.htm', 
        'http://www.wolfbane.com/icd/icd10h.htm')


for(i in 1:length(URL)){
	# Get the data from each URL as a vector of strings 
	# (each vector element is a text line in the URL)
	aux <- read_html(URL[i]) %>% 
  		html_node("pre") %>% 
	  	html_text() %>% 
  		strsplit("\n",fixed = TRUE)

	# Tranform the vector into a list, to facilitate the processing
	aux <- aux[[1]][ 3:length(aux[[1]]) ]	# The first two lines are empty
	attr(aux, "dim") <- c(length(aux),1)	# Transform into matrix
	colnames(aux) <- "text"				# Transform into list

	# Collect the data from the different URL's into a single object raw_icd10
	if(exists("raw_icd10")){
		raw_icd10 <- rbind(raw_icd10, aux)
	} else{
		raw_icd10 <- aux
	}
}


# The object raw_icd10 requires some processing in order to 
# extract the codes and descriptions for each disease in a 
# format that can be later compared to the records of mortality.

# 1) separate...: split each text line into a code and a description. Split at digit-plus-space or digit-parenthesis-space.
# 2) filter...: remove lines that contain text only.
# 3) mutate...: remove spaces that may still remain around the codes and descriptions.
# 4) separate...: this is required for condensed-list codes which have the range of specific codes at the right of the description (see http://www.wolfbane.com/icd/icd10a1.htm).
# 5) select...: we do not need the range of specific codes for each condensed-list code.
# 6) mutate+gsub: eliminate dots and dashes in the codes.

icd10 <- as_tibble(raw_icd10) %>%
	separate(text, into = c("Code", "Cause"), sep = "(?<=[:digit:] )|(?<=[:digit:]\\) )", extra = "merge") %>% 
	mutate(Code = str_trim(Code, side = "both"), Cause = str_trim(Cause, side = "both")) %>%
	separate(Cause, into = c("Cause", "Other"), sep = "(  )", extra = "merge") %>%
	select(-Other) %>%
	mutate(Code = gsub(".", "", Code, fixed = TRUE)) %>%
	mutate(Code = gsub("-", "", Code, fixed = TRUE))


# 4-character codes: these codes are specific cases of 3-character codes. For example:
# C15   Malignant neoplasm of oesophagus	(3-char code)
# C15.0       Cervical part of oesophagus (4-char code)
# C15.1       Thoracic part of oesophagus (4-char code)
# ...
# As observed in this example, the 4-char codes don't always have a complete description of the disease.
# To make each record in icd10 self-sufficient, the description in the 3-char code will be replicated to each of its 4-char codes as well.

for(i in 1:nrow(icd10)){
	# If 4 characters (1 letter plus 3 digits)
	if(str_detect(icd10$Code[i],"[:alpha:][:digit:][:digit:][:digit:]")){
		# Add more general description at the start of Cause
		other = icd10$Cause[icd10$Code == substr(icd10$Code[i],1,3)] # Get the description of the corresponding 3-char code
		icd10$Cause[i] = paste(other,icd10$Cause[i], sep=" - ")	 # Paste the general description at the start of each 4-char record.
	}
}

# Important notes:
# 1) codes starting with CH and UE (Portugal) are still not included in icd10.
# 2) the following codes in the mortality dataset cannot be found in icd10 
# (31 out of 10660 different codes in the mortality database, not counting the codes CH and UE):
# "1000" "AAA"  "C809" "F700" "F701" "F708" "F709" "F710" "F711" "F719" 
# "F720" "F721" "F728" "F729" "F730" "F731" "F738" "F739" "F780" "F781" 
# "F788" "F789" "F790" "F791" "F798" "F799" "I272" "K859" "L899" "M726" "Q315"
# 3) this treatment of the codes is valid only for the icd10 codes. Other codes (icd7, icd8 and icd9) require a different treatment because their format varies substantially.

```


### Keep the **neoplasm** data for ICD10 parts 1 

**To do:** Tranform the data into long format for the number of deaths

**Hints:** Use `gather()`, `filter()`, `str_to_upper()` (or an case insensitive regex to select neoplasm data). 

```{r keepNeoplasm, echo=TRUE}
library(dplyr)
library(tidyr)

# First, I'll change the names of columns Deaths... . It is much easier to do this now than after turning these columns into a new feature. Also this will allow later merging with the population dataset.
morticd10_long <- morticd10_part1
colnames(morticd10_long) <- str_replace(colnames(morticd10_long),"Deaths","")

# Step 1: remove unnecessary columns Admin1, SubDiv, IM_Frmat, IM_Age1,...,IM_Age4 (note thate IM_Deaths... were changed to IM_... by the previous command).
# Step 2: transform to long format: age groups are not variables, but different values of the variable age. 
morticd10_long <- morticd10_long %>%
        select(-Admin1,-SubDiv,-IM_Frmat,-(IM_1:IM_4)) %>%
        gather("AgeGroup", "Deaths", (`1`:`26`)) 

rm(morticd10_part1)
```



### Transform the population data in long format


```{r popLong, echo=TRUE}

# First, I'll change the names of columns Pop... . It is much easier to do this now than after turning these columns into a new feature. Also this will allow later merging with the mortality dataset.
pop_long <- pop
colnames(pop_long) <- str_replace(colnames(pop_long),"Pop","")

# Step 1: remove unnecessary columns Admin1, SubDiv, Lb.
# Step 2: transform to long format: age groups are not variables, but different values of the variable age.
pop_long <- pop_long   %>%
        select(-Admin1, -SubDiv, -Lb) %>%
        gather("AgeGroup", "Population", `1`:`26`)

rm(pop)
```


### Merging the population data to the neoplasm mortality data

  * Care has to be taken with respect to the format of the age groups. 
  * Combine the youngest age groups into a new 0-4 years. 
  * Consider adding an **all** age group category (**not urgent**)
  * Calculate the mortality rate per 100,000 population in a new variable
    + rate = (100000*deaths)/pop

```{r mortalityPop, echo=TRUE}

mortality_pop <- morticd10_long %>%
        	left_join(pop_long, 
		by = c("Country"="Country", 
			"Year"="Year", 
			"Sex"="Sex", 
			"Frmat"="Frmat", 
			"AgeGroup" = "AgeGroup" )) 

# Note 1: it seems like the dataset pop_long does not contain info for certain years and certain countries.
# Note 2: the pop dataset is intended to allow calculating death rates. In that case, does it make sense to keep the population for each age group or should we just keep the population for each set (country, sex, year)?
# Note 3: the mortality database contains entries with NA deaths. It is important to clean both the mortality and the population datasets before merging.
# Note 4: beware that this merging can slow down the computer. The problem is that both morticd10_long and mortality_pop take a big portion of RAM.
```



## Write a function that will extract data based on country, cause of death, year and sex
  * If any of these are left blank then all the data is extracted
  * A data frame containing the requested data should be returned.

```{r}
getWHOdata <- function(country, causes, year, sex){
  # country - Vector of countries to be filtered
  #   cause - Vector of causes of death to be filtered
  #    year - Vector of years to be filtered
  #     sex - Vector of gender to be filtered

}


```


### Next steps

Once the above steps have been completed, the next steps will be defined. They will include:
  * Sumamry graphs
  * Summary tables
  


